Model Recognition as Planning

### 2018-11-20 doc/icaps-19/model-recognition.tex
final version

% Commented for blind submission
\author{Diego Aineto \and Sergio Jim\'enez\and Eva Onaindia\\
{\scriptsize Departamento de Sistemas Inform\'aticos y Computaci\'on}\\
{\scriptsize Universitat Polit\`ecnica de Val\`encia.}\\
{\scriptsize Camino de Vera s/n. 46022 Valencia, Spain}\\
{\scriptsize \{dieaigar,serjice,onaindia\}@dsic.upv.es}
\And Miquel Ram\'irez\\
{\scriptsize School of Computing and Information Systems}\\
{\scriptsize The University of Melbourne}\\
{\scriptsize Melbourne, Victoria. Australia}\\
{\scriptsize miquel.ramirez@nimelb.edu.au}}

@@

A growing interest in addressing the task of recognizing the mission (plans and goals) of a planning agent from a set of given observations has recently emerged. Approaches based on rule-based systems, parsing, graph-covering or Bayesian nets have been proposed for the plan/goal recognition tasks~\cite{sukthankar14}. {\em Plan recognition as planning} is the model-based approach for this task that leverages the action model of the observed agent to compute its most likely goal and predict its future actions~\cite{ramirez2009plan,ramirez2012plan}. While action models are learnable from observations of the agent~\cite{yang2007learning,cresswell2013acquiring,aineto2018learning}, real-world applications developing in a particular context usually dispose of approximate behavioral models of the involved agents and there is rarely a need to learn a model from scratch. A more relevant task in these scenarios is thereby to predict the behavior of an agent from a set of given models.

@@

Figure~\ref{fig:grid-example} illustrates a {\em partial observation} of a five-state trajectory \{{\tt\scriptsize<(xcoord 2)(ycoord 1)>, <(xcoord 3)(ycoord 1)>, <(xcoord 4)(ycoord 1)>, <(xcoord 4)(ycoord 2)>, <(xcoord 3)(ycoord 2)>}\}. Note that this observation only contains fluents of the type {\tt\small (xcoord ?v)} and {\tt\small (ycoord ?v)}, and the value of the remaining fluents, namely {\tt\small (next ?v1 ?v2)}, {\tt\small (q0)} and {\tt\small (q1)}, is not observable in any of the five states.

@@

Editing/learning the action schemata of $\mathcal{M}$ in the AJO approach is addressed converting the task into a classical planning problem, which is later solved with a planner. The intuition behind this compilation is that a solution plan to the problem is a sequence of: (a) \emph{edit actions} on the schemata of $\mathcal{M}$ to build $\mathcal{M'}$ and (b) \emph{validate actions} that apply $\mathcal{M'}$ in $\mathcal{O(\tau)}$. The adaptation of this compilation scheme for solving $\delta^o(\mathcal{M},\mathcal{O})$ results in a planning problem $P'=\tup{F',A',I',G'}$ whose objective is to determine the preconditions and effects that need to be added or deleted to the action schemata of $\mathcal{M}$ so as to satisfy $\mathcal{O}$. The accomplishment of this task requires therefore a propositional encoding of the components of the action schemata. Specifically:

@@

Figure \ref{fig:blocks_acc} shows the {\em accuracy} of our approach when identifying failures over 30 different observations. The positive trend seen in the figure proves that the more complete the observations are the more accurate our estimation becomes.

@@

In this experiment, the planning models of $M$ represent different navigation models for $N\times N$ grids. In particular, we adopt a {\em classical navigation model} with actions {\tt\small up, down, right, left} (to move one cell in each of these four directions) and extend it to define 8 different navigation policies with respect to these additional state variables {\tt\small \{(q0), (q1), (min ?v), (max ?v)\}}. An example of a navigation policy is shown in Figure~\ref{fig:model-example}. This policy allows to move right when $q_0$ holds while it allows to move left when $q_1$ holds, producing a zigzag pattern when visiting all the cells of $N\times N$ grids.

@@

On the other hand, we generated observations from 8 different trajectories, one for each of the 8 previous planning models. The trajectories depict paths followed by a navigation policy to solve the planning problem of {\em visiting all} cells in a $5\times 5$ grid. As in the previous experiments, observations contain no actions and here, for each observed state, only the values of the fluents encoding the {\em x} and {\em y coordinates} of the agent are known (i.e. $\Phi =$\{{\small\tt xcoord, ycoord}\}).

@@

\section{Related work}
\label{sec:related}
An interesting research line related with {\em model recognition} is {\em model reconciliation}~\cite{ChakrabortiSZK17}. In {\em model reconciliation} model edition is used to conform the PDDL models of two different agents, or the model of an agent with a given {\em annotated model}~\cite{sreedharan2018handling}, with respect to a fully observed {\em optimal plan} computed using one of the models. {\em Model recognition} conforms every input model $\mathcal{M}$ in the $M$ set with another model $\mathcal{M}'$ that is consistent with a partial observation of a plan execution. Note, however, that in {\em model recognition} $\mathcal{M}'$ is not given as input but instead is computed for every $\mathcal{M}\in M$. Further, {\em model recognition as planning} shows that the probability of an input planning model to explain a partial observation of a plan execution (that contains no information of the applied actions and where intermediate states may be incomplete or even missing) can be estimated with an off-the-shelf classical planner.

Once the planning model of the observed agent is recognized, then the model-based machinery for automated planning becomes fully applicable. This includes current approaches for {\em goal recognition}~\cite{ramirez2012plan}, {\em goal recognition design}~\cite{KerenGK14} and {\em counter-planning}~\cite{PozancoEFB18} that require to have in advance a fixed model of the observed agent. Last but not least, our approach allows predicting the future actions of an agent whose behavior is not governed by a deliberative planning process aimed to achieve some goals. Our experiments show that {\em model recognition} is suitable for anticipating the actions of agents that are ruled by the execution of fixed programs or finite state machines, such as a {\em regular automata}.

@@

\section{Conclusions}
\label{sec:conclussions}
This paper formalized the {\em model recognition} task and proposed, {\em model recognition as planning}, a method built on top of off-the-shelf classical planning algorithms to estimate the probability of a \strips\ model to produce a partial observation of a plan execution. The paper shows the effectiveness of {\em model recognition as planning} in a set of \strips\ models encoding different kinds of {\em automata}. {\em Model recognition as planning} succeeds to identify the executed automata despite the internal machine state or actual applied transitions, are unobserved.

>>>>

% Commented for blind submission
%\author{Diego Aineto \and Sergio Jim\'enez\and Eva Onaindia\\
%{\scriptsize Departamento de Sistemas Inform\'aticos y Computaci\'on}\\
%{\scriptsize Universitat Polit\`ecnica de Val\`encia.}\\
%{\scriptsize Camino de Vera s/n. 46022 Valencia, Spain}\\
%{\scriptsize \{dieaigar,serjice,onaindia\}@dsic.upv.es}
%\And Miquel Ram\'irez\\
%{\scriptsize School of Computing and Information Systems}\\
%{\scriptsize The University of Melbourne}\\
%{\scriptsize Melbourne, Victoria. Australia}\\
%{\scriptsize miquel.ramirez@nimelb.edu.au}}

@@

A growing interest in addressing the task of recognizing the mission (plans and goals) of a planning agent from a set of given observations has recently emerged. Approaches based on rule-based systems, parsing, graph-covering or Bayesian nets have been proposed for the plan/goal recognition tasks~\cite{sukthankar14}. {\em Plan recognition as planning} is the model-based approach for this task that leverages the action model of the observed agent to compute its most likely goal and predict its future actions~\cite{ramirez2009plan,ramirez2012plan}. While action models are learnable from observations of the agent~\cite{yang2007learning,cresswell2013acquiring,aineto2018learning}, in real-world applications developing in a particular context, approximate behavioral models of the involved agents (manually-introduced or automatically-learned) are already available, so there is rarely a need to learn a model from scratch. A more relevant and applicable task in these scenarios is thereby to predict the behavior of an agent from a set of given models.

@@

Figure~\ref{fig:grid-example} illustrates a partial observation of a six-state trajectory \{{\tt\scriptsize<(xcoord 0)(ycoord 0)>, <(xcoord 1)(ycoord 0)>, <(xcoord 2)(ycoord 0)>, <(xcoord 3)(ycoord 0)>, <(xcoord 3)(ycoord 1)>, <(xcoord 2)(ycoord 1)>}\}. This observation only contains fluents of the predicates {\tt\small (xcoord ?v)} and {\tt\small (ycoord ?v)}, and the value of the remaining fluents, corresponding to predicates {\tt\small (next ?v1 ?v2)}, {\tt\small (q0)} and {\tt\small (q1)}, is unobservable in the six states.

@@

Editing/learning the action schemata of $\mathcal{M}$ in the AJO approach is addressed converting the task into a classical planning problem, which is later solved with a planner. The intuition behind this compilation is that a solution plan to the problem is a sequence of: (a) \emph{edit actions} on the schemata of $\mathcal{M}$ to build $\mathcal{M'}$ and (b) \emph{validate actions} that apply $\mathcal{M'}$ in $\mathcal{O(\tau)}$. The adaptation of this compilation scheme for solving $\delta^o(\mathcal{M},\mathcal{O})$ results in a planning problem $P'=\tup{F',A',I',G'}$ whose objective is to determine the preconditions and effects that need to be added or deleted to the action schemata of $\mathcal{M}$ so as to satisfy $\mathcal{O}$. The accomplishment of this task requires therefore a propositional encoding of the components of the action schemata:

@@

Figure \ref{fig:blocks_acc} shows the {\em accuracy} of our approach when identifying failures over 30 different observations across different degrees of observability. The positive trend seen in the figure proves that the more complete the observations are, the more accurate our estimation becomes.

@@

The planning models of $M$ represent different navigation models for $N\times N$ grids. In particular, we adopt a {\em classical navigation model} with actions {\tt\small up, down, right, left} (to move one cell in each of these four directions) and extend it to define 8 different navigation policies with respect to these additional state variables {\tt\small \{(q0), (q1), (min ?v), (max ?v)\}}. An example of a navigation policy is shown in Figure~\ref{fig:model-example}. This policy allows to move right when $q_0$ holds while it allows to move left when $q_1$ holds, producing a zigzag pattern when visiting all the cells of $N\times N$ grids.

@@

On the other hand, we generated observations from 8 different trajectories, one for each of the 8 previous planning models. The trajectories depict paths followed by a navigation policy to solve the planning problem of {\em visiting all} cells in a $5\times 5$ grid. As in the previous experiments, observations contain no actions and here, for each observed state, only the values of the fluents encoding the {\em x} and {\em y coordinates} of the agent are known; i.e. $\Phi$ comprises all the fluents instantiated with predicates {\tt\small (xcoord ?v)} and {\tt\small (ycoord ?v)}.

@@

\section{Conclusions and Related work}
\label{sec:related}

This paper formalizes the task of {\em model recognition} and introduces a novel method that estimates the probability of a \strips\ model to explain a partial observation of a plan execution. Our {\em model recognition as planning} approach builds on top of off-the-shelf classical planning algorithms and it is robust to missing actions and incomplete or missing intermediate states in the observation. Once the planning model of the observed agent is recognized, the model-based machinery for automated planning becomes fully applicable for other recognition tasks like {\em goal recognition}~\cite{ramirez2012plan}, {\em goal recognition design}~\cite{KerenGK14} or {\em counter-planning}~\cite{PozancoEFB18}, which require beforehand a model of the observed agent.

We show the effectiveness of our approach in three experiments that encode a set of different \strips\ models. 
The first experiment is a proof-of-concept \emph{classification task} that perfectly recognizes the regular automata of a given string. Additionally, the two assumptions underpinning our formal theory are experimentally proved in the second experiment by varying the degree of observability. Finally, the robustness of the approach is shown in the third experiment where actions and intermediates states are unobservable. A meaningful lesson from the experiments is that our approach allows us to predict the future actions of agents whose behavior is ruled by a  model other than a deliberative planning model aimed at achieving some goals; this includes, for instance, models of fixed programs or finite state machines (regular automata).

An interesting research line related to {\em model recognition} is {\em model reconciliation}~\cite{ChakrabortiSZK17}, in which model edition is used to conform the PDDL models of two agents, or the model of an agent with a given {\em annotated model}~\cite{sreedharan2018handling}, with respect to a fully observed optimal plan computed with one of the two models. {\em Model recognition}, however, conforms every input model $\mathcal{M}$ with another model $\mathcal{M}'$ that is not given as input but instead computed for every $\mathcal{M}$, and which is consistent with a partial observation of a plan execution.

@@
